---
title: "K size tests"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Set-up

```{r}
datafile <- file.path(here::here(), "data", "SEAS-monthly-sst.RData")
load(datafile)
```

Packages
```{r}
library(factoextra)
```

Set up the data and data features. 
```{r}
Data_clean <- out$dat.clean
Data_dirty <- out$dat
pos_loc <- out$pos.loc
```

Data clean and meta data. Remove the lat/lon rows and change to Celcius.
```{r}
lats <- Data_dirty[1,]
lons <- Data_dirty[2,]
asp <- c(table(lons)[1], table(lats)[1]) # lon (x) - lats (y)
bb.box <- c(min(lons), max(lons), min(lats), max(lats))
```

Create matrices for `kmeans()`
```{r}
X <- Data_clean[c(-1,-2),] - 273.15
X_norm <- t(scale(t(X), scale=FALSE))
```

Set variables for `kmeans()`
For the pilot analysis, set K = 12. Some relatively big to try to capture the variability.
```{r}
iter.max <- 25
nstart <- 100
```

## Heat maps

Function for the analyses.
```{r echo=FALSE}
findK <- function(X, nstart=25, iter.max=100, n_K=15){
cost <- rep(0, n_K-1)
for(k in  1:(n_K-1)) 
  cost[k] <- kmeans(X, k, iter.max=iter.max, nstart=nstart)$tot.withinss
plot(1:(n_K-1), cost, xlab="K", ylab="total within ss", type="b")
}
kheatmap <- function(n_K, X_norm, Data_dirty, main.n=5, iter.max=25, nstart=100){
  set.seed(1221966)
  out_kmeans <- kmeans(X_norm, n_K, iter.max=iter.max, nstart=nstart)
  centroid_images <- matrix(NA, n_K, ncol(Data_dirty))
  centroid_images[, pos_loc] <- out_kmeans$centers
  rownames(centroid_images) <- paste("Centroid", 1:n_K)
  library(raster)
  img.list <- list()
  img.mat <- centroid_images
  for(i in 1:nrow(img.mat)){
  tmp <- matrix(img.mat[i,], asp[2], asp[1], byrow=TRUE)
  tmp <- tmp[asp[2]:1, ] # lat 7 at bottom not top
  tmp <- raster(tmp)
  extent(tmp) <- bb.box
  img.list[[i]] <- tmp
  }
CC <- out_kmeans$centers
labs <- paste0("C", 1:n_K)
rownames(CC) <- labs
pal <- colorRamps::matlab.like(100)
dend <- dendIMG(CC, img.list, type="original", pal=pal, scale.same=TRUE, lab.extra="none")
rect.dendrogram( dend, k=main.n, lty = 5, lwd = 0, col=rgb(0.1, 0.2, 0.4, 0.1) ) 
ph <- myheatmap(dend, out_kmeans, main.n=main.n, sub.n=n_K, ramp="var", ramp.type=2)
plot(ph)
return(ph)
}
```

```{r echo=FALSE}
fviz1 <- function(n_K, X_norm, Data_dirty, main.n=4, iter.max=25, nstart=100, type=1){
  set.seed(1221966)
  out_kmeans <- kmeans(X_norm, n_K, iter.max=iter.max, nstart=nstart)
CC <- out_kmeans$centers
labs <- paste0("C", 1:n_K)
rownames(CC) <- labs
if(type==2) fviz_nbclust(CC, kmeans, method = "silhouette",k.max=n_K-1)
if(type==1) fviz_nbclust(CC, kmeans, method = "wss",k.max=n_K-1)
if(type==3){
  gap_stat <- cluster::clusGap(CC, FUN = kmeans, nstart = 25,
                    K.max = n_K-1, B = 50)
  fviz_gap_stat(gap_stat)
}
if(type==4) NbClust::NbClust(CC, distance = "euclidean", min.nc=2, max.nc=4, 
            method = "ward.D", index = "all")
}
```


```{r echo=FALSE}
source(file.path(here::here(), "R", "myheatmap.R"))
source(file.path(here::here(), "R", "desat.R"))
source(file.path(here::here(), "R", "dendIMG.R"))
```

### K=4

```{r}
p4 <- kheatmap(4, X_norm, Data_dirty, main.n=4)
```

### K=6

```{r}
p6 <- kheatmap(6, X_norm, Data_dirty)
```

### K=8

```{r}
p8 <- kheatmap(8, X_norm, Data_dirty)
```

### K=10

```{r}
p10 <- kheatmap(10, X_norm, Data_dirty)
```

### K=12

```{r k12}
p12 <- kheatmap(12, X_norm, Data_dirty)
```

### K=14

```{r k14}
p14 <- kheatmap(14, X_norm, Data_dirty)
```

### K=16

```{r k16}
p16 <- kheatmap(16, X_norm, Data_dirty)
```

### K=18

```{r k18}
p18 <- kheatmap(18, X_norm, Data_dirty)
```

### K=20

```{r k20}
p20 <- kheatmap(20, X_norm, Data_dirty)
```

### K=22

```{r k22}
p22 <- kheatmap(22, X_norm, Data_dirty)
```

### K=25

```{r k25}
p25 <- kheatmap(25, X_norm, Data_dirty, main.n=6)
```

### K=30

```{r k30}
p30 <- kheatmap(30, X_norm, Data_dirty, main.n=6)
```

## All together

It appears that at least K=14 is necessary to capture the May, to June, to July switches in patterns. The red line is June.
```{r echo=FALSE}
library(gridExtra)
thm <- theme(legend.position = "none", axis.title = element_blank(), axis.text = element_blank())
add <- geom_hline(yintercept=6, col="red", alpha=0.3)
grid.arrange(p4+thm+ggtitle("K = 4")+add,
             p6+thm+ggtitle("K = 6")+add, 
             p8+thm+ggtitle("K = 8")+add, 
             p10+thm+ggtitle("K = 10")+add,
             p12+thm+ggtitle("K = 12")+add,
             p14+thm+ggtitle("K = 14")+add,
             p16+thm+ggtitle("K = 16")+add,
             p18+thm+ggtitle("K = 18")+add,
             p20+thm+ggtitle("K = 20")+add,
             p22+thm+ggtitle("K = 22")+add,
             p25+thm+ggtitle("K = 25")+add,
             p30+thm+ggtitle("K = 30")+add)
```


