---
title: "Using unsupervised image classification to study changes in seasonal upwelling in the Southeast Arabian Sea"
subtitle: "Elizabeth Eli Holmes, NWFSC/UW SAFS"
date: "9 Apr 2021 SAFS Quantative Seminar"
output:
  ioslides_presentation:
    css: talk.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Topics

* Motivation for this study
* Upwelling patterns
* Common types of unsupervised image classification
  * PCA
  * K-means
  * Hierarchical clustering
* Seasonal patterns of upwelling revealed by image classification
* Changes in upwelling in the SEAS

## Unsupervised image classification

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

